### 2015-05-11 papers/histogram/thesis/thesis.tex
minor changes and an acknowledgments statement

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Metropolis Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low (i.e. liquid-state) temperatures. Previous
works have developed several generic Monte Carlo histogram methods for
collecting statistics on low energy system states, but little or no
literature exists on these methods' systematic comparison, as well as
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model). We implement the Wang-Landau
method, transition matrix Monte Carlo, as well as two methods of our
own. The first of our methods is novel, while the second is a hybrid
of two published methods (transition matrix Monte Carlo and the
optimized ensemble).

We find that Wang-Landau potentially performs better than other
histogram methods, but fails catastrophically without a predefined
energy range for the square-well fluid. Our first method gives results
comparable to successful Wang-Landau simulations, but can have
anomalous cases with large error. The transition matrix Monte Carlo
consistently yields results whose quality is marginally worse than
functional Wang-Landau. Finally, our implementation of the optimized
ensemble using a transition matrix results in worse performance over
the straightforward transition matrix Monte Carlo method.

@@

\subsection{Monte Carlo fluid simulations}
\label{sec:mc_sim}

While model systems are powerful tools for understanding complex
physical systems, they do not themselves exist in the real
world. Consequently, direct experimental tests of theories for model
system (e.g. square-well density functional theories) are not
possible. For this reason, model system theories are commonly tested
against Monte Carlo simulations. Proper implementation of Monte Carlo
methods to study completely characterized systems ensures that
statistical results from simulations converge on the exact properties
of the simulated system in the infinite simulation time
limit. Furthermore, uncertainties in quantities computed via Monte
Carlo simulations are typically well-defined, monotonically decreasing
functions of simulation time, allowing one to run simulations to the
desired level of accuracy.

@@

Algorithm \ref{alg:metropolis} provides a sketch of unbiased,
Metropolis Monte Carlo fluid simulations. Such an algorithm is
``unbiased'' in the sense that it collects statistics (i.e. data) on
all valid system configurations with equal probability. Statistics
whose collection time scales as $\mathcal O\p{1}$ with system size,
meaning that increasing the number of spheres $N$ in the simulation
does not affect collection time, can be collected after every move
(defined in the algorithm). Statistics whose collection time scales as
$\mathcal O\p{N}$, meaning that doubling $N$ doubles collection time,
can be collected after each iteration. In general, collection with
$\mathcal O\sp{\chi\p{N}}$ time scaling should not occur more often
than once every $\chi\p{N}$ moves, where $\chi\p{N}$ may be an
arbitrary function of $N$ void of constant prefactors, e.g. $N\log N$,
or $2^N$. These collection rules ensure that scaling up simulations
does not cause them to asymptotically spend all computation time only
collecting statistics, or only simulating the fluid. Collected
statistics are used to find thermodynamic properties of the simulated
fluid.

@@

As well as the density of states, the transition matrix $T\p{E_i,E_j}$
determines the local diffusivity $\alpha\p{E}$ for any given weights
$w\p{E}$. Denoting by $P_b\p{E_i\to E_j}$ the actual, biased
probability of transitioning from $E_i$ to $E_j$ in a single Monte
Carlo move, we can say that
\begin{align}
  P_b\p{E_i\to E_f}=P\p{E_i\to E_f}P_m\p{E_i\to E_f}
  =T_{ji}\max\set{\f{w\p{E_f}}{w\p{E_i}},1},
\end{align}
where $P\p{E_i\to E_j}$ is the probability of attempting the move
$E_i\to E_j$ and $P_m\p{E_i\to E_j}$ is the probability of accepting
that move, should it occur. In terms of a transition histogram $\tilde
T\p{E,\Delta E}$ with $E=E_i$ and $\Delta E=E_f-E_i$,
\begin{align}
  P_b\p{E\to E+\Delta E} =\f{\tilde T_d\p{E,\Delta E}w\p{E+\Delta
      E}/w\p{E}} {\sum_{\Delta E'}\tilde T_d\p{E,\Delta
      E'}w\p{E+\Delta E'}/w\p{E}}.
  \label{eq:actual_transition_prob}
\end{align}
We now define
\begin{align}
  \bk{\Delta E}_E=\sum_{\Delta E}\Delta E P_b\p{E\to E+\Delta E}
  =\f{\sum_{\Delta E}\Delta E~\tilde T_d\p{E,\Delta E}
    \max\set{w\p{E+\Delta E}/w\p{E},1}} {\sum_{\Delta E'}\tilde
    T_d\p{E,\Delta E'} \max\set{w\p{E+\Delta E'}/w\p{E},1}}
  \label{eq:<de>}
\end{align}
and
\begin{align}
  \bk{\Delta E^2}_E=\sum_{\Delta E}\Delta E^2 P_b\p{E\to E+\Delta E}
  =\f{\sum_{\Delta E}\Delta E~\tilde T_d\p{E,\Delta E}
    \max\set{w\p{E+\Delta E}/w\p{E},1}} {\sum_{\Delta E'}\tilde
    T_d\p{E,\Delta E'} \max\set{w\p{E+\Delta E'}/w\p{E},1}},
  \label{eq:<de^2>}
\end{align}
in terms of which the diffusivity $\alpha\p{E}$ is
\begin{align}
  \alpha\p{E}=\bk{\Delta E^2}_E -\bk{\Delta E}_E^2.
  \label{eq:microscopic_diffusitivy}
\end{align}
Diffusivity $\alpha\p{E}$ is thus the variance of the change in energy
$\Delta E$ from as single move out of $E$. Reconciling this definition
with that we provide in (\ref{eq:walker_current_definition}) is
considered outside the scope of this paper.

@@

>>>>

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Metropolis Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low (i.e. liquid-state) temperatures. Previous
works have developed several generic Monte Carlo histogram methods for
collecting statistics on low energy system states, but little or no
literature exists on these methods' systematic comparison, as well as
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model). We implement the Wang-Landau
method, transition matrix Monte Carlo, as well as two methods of our
own. The first of our methods, the simple flat method, is novel, while
the second is a hybrid of two published methods (transition matrix
Monte Carlo and the optimized ensemble).

We find that Wang-Landau potentially performs better than other
histogram methods, but fails catastrophically without a predefined
energy range for the square-well fluid. The simple flat method and
transition matrix Monte Carlo give results comparable to successful
Wang-Landau simulations, but simple flat has some anomalous cases with
large errors. Finally, our implementation of the optimized ensemble
using a transition matrix results in worse performance over the
straightforward transition matrix Monte Carlo method.

@@

\subsection{Monte Carlo fluid simulation}
\label{sec:mc_sim}

Model systems such as the square-well fluid are powerful tools for
understanding complex physical systems, but they do not themselves
exist in the real world. Consequently, direct experimental tests of
theories for model system (e.g. square-well density functional
theories) are not possible. For this reason, model system theories are
commonly tested against Monte Carlo simulations. Proper implementation
of Monte Carlo methods to study completely characterized systems
ensures that statistical results from simulations converge on the
exact properties of the simulated system in the infinite simulation
time limit. Furthermore, uncertainties in quantities computed via
Monte Carlo simulations are typically well-defined, monotonically
decreasing functions of simulation time, allowing one to run
simulations to the desired level of accuracy.

@@

Algorithm \ref{alg:metropolis} provides a sketch of unbiased,
Metropolis Monte Carlo fluid simulation. Such an algorithm is
``unbiased'' in the sense that it collects statistics (i.e. data) on
all valid system configurations with equal probability. Statistics
whose collection time scales as $\mathcal O\p{1}$ with system size,
meaning that increasing the number of spheres $N$ in the simulation
does not affect collection time, can be collected after every move
(defined in the algorithm). Statistics whose collection time scales as
$\mathcal O\p{N}$, meaning that doubling $N$ doubles collection time,
can be collected after each iteration. In general, collection with
$\mathcal O\sp{\chi\p{N}}$ time scaling should not occur more often
than once every $\chi\p{N}$ moves, where $\chi\p{N}$ may be an
arbitrary function of $N$ void of constant prefactors, e.g. $N\log N$,
or $2^N$. These collection rules ensure that scaling up simulations
does not cause them to asymptotically spend all computation time only
collecting statistics, or only simulating the fluid. Collected
statistics are used to find thermodynamic properties of the simulated
fluid.

@@

As well as the density of states, the transition matrix $T\p{E_i,E_j}$
determines the local diffusivity $\alpha\p{E}$ for any given weights
$w\p{E}$. Denoting by $P_b\p{E_i\to E_j}$ the actual, biased
probability of transitioning from $E_i$ to $E_j$ in a single Monte
Carlo move, we can say that
\begin{align}
  P_b\p{E_i\to E_f}=P\p{E_i\to E_f}P_m\p{E_i\to E_f}
  =T_{ji}\max\set{\f{w\p{E_f}}{w\p{E_i}},1},
\end{align}
where $P\p{E_i\to E_j}$ is the probability of attempting the move
$E_i\to E_j$ and $P_m\p{E_i\to E_j}$ is the probability of accepting
that move, should it occur. In terms of a transition histogram $\tilde
T_d\p{E,\Delta E}$ with $E=E_i$ and $\Delta E=E_f-E_i$,
\begin{align}
  P_b\p{E\to E+\Delta E} =\f{\tilde T_d\p{E,\Delta E}w\p{E+\Delta
      E}/w\p{E}} {\sum_{\Delta E'}\tilde T_d\p{E,\Delta
      E'}w\p{E+\Delta E'}/w\p{E}}.
  \label{eq:actual_transition_prob}
\end{align}
We now define
\begin{align}
  \bk{\Delta E}_E=\sum_{\Delta E}\Delta E P_b\p{E\to E+\Delta E}
  =\f{\sum_{\Delta E}\Delta E~\tilde T_d\p{E,\Delta E}
    \max\set{w\p{E+\Delta E}/w\p{E},1}} {\sum_{\Delta E'}\tilde
    T_d\p{E,\Delta E'} \max\set{w\p{E+\Delta E'}/w\p{E},1}}
  \label{eq:<de>}
\end{align}
and
\begin{align}
  \bk{\Delta E^2}_E=\sum_{\Delta E}\Delta E^2 P_b\p{E\to E+\Delta E}
  =\f{\sum_{\Delta E}\Delta E~\tilde T_d\p{E,\Delta E}
    \max\set{w\p{E+\Delta E}/w\p{E},1}} {\sum_{\Delta E'}\tilde
    T_d\p{E,\Delta E'} \max\set{w\p{E+\Delta E'}/w\p{E},1}},
  \label{eq:<de^2>}
\end{align}
in terms of which the diffusivity $\alpha\p{E}$ is
\begin{align}
  \alpha\p{E}=\bk{\Delta E^2}_E -\bk{\Delta E}_E^2.
  \label{eq:microscopic_diffusitivy}
\end{align}
The diffusivity $\alpha\p{E}$ is thus the variance of the change in
energy $\Delta E$ from as single move out of $E$. Reconciling this
definition with that we provide in
(\ref{eq:walker_current_definition}) is considered outside the scope
of this paper. \fixme{say something more convincing? cite?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Acknowledgments}
\label{sec:acknowledgements}

I wish to thank, first and foremost, my advisor Dr. David Roundy, for
mentoring me and providing me a wealth of knowledge and expertise in
various subjects of physics, numerical methods, and programming
nuances. I would also like thank Paho Lurie-Gregg, with whom I
frequently exchange and discuss codes and ideas, and whose hard
polyhedra Monte Carlo simulations I used as the template for my own
codes. Lastly, I would like to thank thank Eva Morgun, who prompted me
to {\it act} on my textbook enthusiasm for physics, and get involved
in the fun and interesting world of scientific research.
